Wigglyhedra

### 2024-06-28 draft/wigglyhedra.tex
Updated figure and writeup about arcs.

\pagebreak
\para{Categorical interpretation}
Each wiggly arc on \(n+2\) points corresponds to an object of an abelian category, such that morphisms and extensions between objects in the category are captured by minimal intersection numbers between the arcs.
In fact, this correspondence holds for more general arcs, which then correspond to objects of a larger triangulated category~\cite{kho.sei:02}.

A general arc can be decomposed into wiggly arcs simply by cutting along U-turns.
We interpret this operation categorically as the decomposition of an object of the triangulated category into its cohomology pieces, from which we deduce the following.
\begin{proposition*}
  Consider the set of wiggly arcs corresponding to the cohomology pieces of an object represented by a general arc.
  This set is contained in a face of the wiggly complex.
\end{proposition*}
Furthermore, we characterise categorically when two objects in the abelian category are compatible, that is, their corresponding wiggly arcs are compatible.
\begin{proposition*}
  Consider two wiggly arcs, and their corresponding objects in the abelian category.
  The wiggly arcs are compatible if and only if there are no extensions between the corresponding objects.
\end{proposition*}

@@

\begin{figure}
\centerline{
	\begin{tabular}{c@{\qquad}c@{\qquad}c}
		Bounded homotopy category 
		&
		&
		\\
		of graded projective modules 
		&
		$\smash{\xrightarrow[\quad\text{combinatorial model}\quad]{}}$
		&
		curves
		\\
		over the zigzag algebra
		&
		&
		\\
		COMPLEX
		&
		&
		\includegraphics[scale=1.3]{curve1}
		\\[.5cm]
		cohomology $\Bigg\downarrow$ filtration
		&
		&
		$\Bigg\downarrow$ cut
		\\[.5cm]
		Heart of the standard $t$-structure
		&
		$\xrightarrow[\quad\text{combinatorial model}\quad]{}$
		&
		wiggly arcs
		\\
		COMPLEX1
		&
		&
		\includegraphics[scale=1.3]{curve2}
		\\
		COMPLEX2
		&
		&
		\includegraphics[scale=1.3]{curve3}		
	\end{tabular}
}
\caption{An illustration of the Khovanov--Seidel curve model and \cref{prop:cohomology-pieces}.}
\label{fig:curveModel}
\end{figure}

@@

Consider \(n+2\) points arranged in a straight horizontal line, and labelled \(0,1, \ldots, n+1\) as in~\cref{sec:combinatorics}.
In addition to the wiggly arcs, consider isotopy classes of curves whose endpoints lie on distinct marked points, and which do not cross themselves or any marked point in their interior.
For convenience, we refer to such an isotopy class as an \defn{arc}.

@@

\begin{proposition}\label{prop:cohomology-pieces}
  Let \(\alpha\) be any arc, and let \(X = X^{\bullet}\) be a chain complex that corresponds to \(\alpha\) under the Khovanov--Seidel reconstruction.
  Recall that for any integer \(j\), the cohomology object \(H^j(X)\) is an object of \(\mathcal{A}_m\), and hence a direct sum of indecomposable objects of \(\mathcal{A}_m\).
  \begin{enumerate}
  \item The object \(X\) is an indecomposable spherical object of \(\mathcal{A}_m\) up to shift if and only if \(\alpha\) is a wiggly arc.
    In this case, \(H^j(X) \neq 0\) for exactly one \(j\), such that \(X \cong H^j(X)[-j]\).
  \item If \(\alpha\) is a general arc, each indecomposable summand of \(H^j(X)\) for any \(j\) is a spherical indecomposable object of \(\mathcal{A}_m\) that arises from a wiggly arc, up to shift.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Let \(X\) be a complex built from an arbitrary (not necessarily wiggly) arc \(\alpha\).
  The results of~\cite{kho.sei:02} imply that \(X\) is spherical and indecomposable.
  Let us now analyse the indecomposable summands of its cohomology objects.

@@

For a general object \(X\) of \(\mathcal{C}_m\), the indecomposable factors of the cohomology filtration can be arbitrary.
However, it turns out that the cohomology filtration is highly constrained if \(X\) is spherical, and these constraints are given by wiggly pseudotriangulations.
\begin{proposition}\label{prop:hn-filtration}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} & \cdots & H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of all possible indecomposable summands of the cohomology objects \(\{H^j(X)\}\) taken without multiplicity.
  Let \(T\) be corresponding set of wiggly arcs, as explained in~\cref{prop:cohomology-pieces}.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of this proposition is an easy consequence of analysing the cohomology filtration from the point of view of arcs on the disk with marked points.
  Let \(X_1, \ldots, X_k\) be the objects in \(T\) without repetition, which arise from wiggly arcs \(\alpha_1, \ldots, \alpha_k\) as explained in the proof of~\cref{prop:cohomology-pieces}.
  Let \(\beta = (i,j,A,B)\) and \(\beta' = (i',j', A',B')\) be two distinct wiggly arcs among \(\{\alpha_1, \ldots,\alpha_k\}\).
  We check that \(\beta\) and \(\beta'\) are compatible in the sense of~\cref{def:compatible}.

  Recall that \(\beta\) and \(\beta'\) are two pieces of the arc \(\alpha\).
  Therefore \(\beta\) and \(\beta'\) cannot cross, because~\(\alpha\) does not cross itself.
  Suppose that \(\beta\) and \(\beta'\) are non-pointed, and without loss of generality, suppose that \(i' = j\).
  This means that the piece \(\beta\) was cut from \(\alpha\) either because \(i' = j\) is an endpoint of \(\alpha\), or because \(\alpha\) loops around the right of the point \(i'\).
  Moreover, if \(i'\) is an endpoint of \(\alpha\), then the portion of \(\alpha\) closest to \(i'\) lies to the left of \(i'\).

@@

  We have four possible combinations:
  \begin{enumerate}
  \item \(i' = j\) is an endpoint such that \(\alpha\) goes left starting at \(i'\), and \(i' = j\) is an endpoint such that \(\alpha\) goes right starting at \(j\);
  \item \(i' = j\) is an endpoint such that \(\alpha\) goes left starting at \(i'\), and \(\alpha\) loops around the left of the point \(j\);
  \item \(\alpha\) loops around the right of the point \(i' = j\), and \(i' = j\) is an endpoint such that \(\alpha\) goes right starting at \(j\); 
  \item \(\alpha\) loops around the right of the point \(i' = j\), and \(\alpha\) loops around the left of the point \(j\).
  \end{enumerate}
  It is clear that none of these possibilities can occur within the arc \(\alpha\), and hence \(\beta\) and \(\beta'\) must be pointed.
  Therefore \(T = \{\alpha_1, \ldots, \alpha_k\}\) is a subset of a wiggly pseudotriangulation.

@@

However, the grading convention in~\cite{kho.sei:02} is not the same as ours --- our grading is by path length, while the grading in~\cite{kho.sei:02} is one for arrows in one direction and zero for arrows in the opposite direction.
So we must modify their argument slightly.
In particular, the table in~\cite[Lemma 3.20]{kho.sei:02} that computes the bigraded intersection numbers will change, and so will the table in~\cite[Lemma 4.12]{kho.sei:02} that computes the Poincar{\'e} polynomials of \(\Hom(P_k,-)\).
  \asilata{TODO: this is very tedious; how much to write it?}
\end{proof}
\begin{remark}
  \cref{prop:ext1} is very similar to the condition of being compatible in a cluster-tilting object of a cluster category (see, \eg~\cite[Cor.~4.3]{bua.mar.rei.ea:06}) or in a $\tau$-tilting object (see, \eg~\cite{AdachiIyamaReiten}).
  The condition in terms of crossings of curves already appears for some specializations (see, \eg~\cite{FominShapiroThurston, PaluPilaudPlamondon-surfaces}).
  However, in \cref{prop:ext1}, it depends on shifting the objects~\(X\) and~\(X'\) so that they both lie in the standard heart, which is an auxiliary step that is not required in cluster-tilting and $\tau$-tilting theory.
\end{remark}
\begin{remark}
  As stated,~\cref{prop:ext1} in combination with~\cref{prop:hn-filtration} is not expected to be true in greater generality.
  For instance, for the zigzag algebra of Dynkin type~\(D_4\), one can find two indecomposable spherical objects with a non-trivial \(\Ext^1\) that nevertheless appear together in the cohomology filtration of another spherical object.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
\end{remark}
\begin{question}
  Suppose we have a collection of spherical objects in the standard heart \(\mathcal{A}_m\) that satisfy the compatibility condition of~\cref{prop:ext1}.
  That is, the \(\Ext^1\) between any pair of objects in this collection is zero.
  \begin{enumerate}
  \item Can we prove properties about a maximal such collection without reference to arcs?
  \item In particular, can we understand the combinatorics of maximal such collections without reference to arcs?
  \item Given a maximal compatible collection of objects, can we always find a larger spherical object such that the factors of the cohomology filtration of that object are precisely shifts of the objects in the given collection?
  \end{enumerate}
\end{question}

@@

%%% Local variables:
%%% TeX-command-extra-options: "-pvc"
%%% End:

>>>>

\pagebreak
\para{Categorical interpretation}
Each wiggly arc on \(n+2\) points corresponds to an object of an abelian category, such that morphisms and extensions between objects in the category are captured by minimal intersection numbers between the arcs.
In fact, this correspondence holds for more general curves, which then correspond to objects of a larger triangulated category~\cite{kho.sei:02}.

A general curve can be decomposed into wiggly arcs simply by cutting along U-turns.
We interpret this operation categorically as the decomposition of an object of the triangulated category into its cohomology pieces, from which we deduce the following.
\begin{proposition*}
  Consider the set of wiggly arcs corresponding to the cohomology pieces of an object represented by a general arc.
  This set is contained in a face of the wiggly complex.
\end{proposition*}
Furthermore, we characterise categorically when two objects in the abelian category are compatible, that is, their corresponding wiggly arcs are compatible.
\begin{proposition*}
  Consider two wiggly arcs, and their corresponding objects in the abelian category.
  The wiggly arcs are compatible if and only if there are no extensions between the corresponding objects.
\end{proposition*}

@@

\begin{figure}
\centerline{
	\begin{tabular}{c@{\qquad}c@{\qquad}c}
		Spherical objects in bounded 
		&
		&
		\\
		homotopy category of graded projective 
		&
		$\smash{\xrightarrow[\quad\text{combinatorial model}\quad]{}}$
		&
		Curves
		\\
		modules over the zigzag algebra
		&
		&
                \\
          \begin{tikzcd}[ampersand replacement=\&, row sep=-0.1cm, column sep=0.5cm]
                                  \&           \& P_3 \langle -1 \rangle\arrow{dr} \& \\
                                  \&           \&                      \& P_2\\
            P_1\langle -5 \rangle\arrow{r}    \& P_1\langle -3 \rangle\arrow{r}   \& P_1 \langle -1 \rangle\arrow{ur} \& 
          \end{tikzcd}
		&
		&
		\includegraphics[scale=1.3]{curve1}
		\\[.5cm]
		{\scriptsize cohomology} $\Bigg\downarrow$ {\scriptsize filtration}
		&
		&
		$\Bigg\downarrow$ {\scriptsize cut}
		\\[.5cm]
		Heart of the standard $t$-structure
		&
		$\xrightarrow[\quad\text{combinatorial model}\quad]{}$
		&
		Wiggly arcs
		\\
          \begin{tikzcd}[ampersand replacement=\&, row sep=-0.1cm, column sep=0.5cm]
            P_1\langle -1 \rangle \arrow{dr}\& \\
            \& P_2 \\
            P_3 \langle -1 \rangle \arrow{ur}
          \end{tikzcd}
		&
          &
		\includegraphics[scale=1.3]{curve2}
          \\
          \(P_1\)

		&
		&
		\includegraphics[scale=1.3]{curve3}		
	\end{tabular}
}
\caption{An illustration of the Khovanov--Seidel curve model and \cref{prop:cohomology-pieces}.}
\label{fig:curveModel}
\end{figure}

@@

Consider \(n+2\) points arranged in a straight horizontal line, and labelled \(0,1, \ldots, n+1\) as in~\cref{sec:combinatorics}.
In addition to the wiggly arcs, consider isotopy classes of curves whose endpoints lie on distinct marked points, and which do not cross themselves or any marked point in their interior.
For convenience, we refer to such an isotopy class as a \defn{curve}.

@@

\begin{proposition}\label{prop:cohomology-pieces}
  Let \(\alpha\) be any curve, and let \(X = X^{\bullet}\) be a chain complex that corresponds to \(\alpha\) under the Khovanov--Seidel reconstruction.
  Recall that for any integer \(j\), the cohomology object \(H^j(X)\) is an object of \(\mathcal{A}_m\), and hence a direct sum of indecomposable objects of \(\mathcal{A}_m\).
  \begin{enumerate}
  \item The object \(X\) is an indecomposable spherical object of \(\mathcal{A}_m\) up to shift if and only if \(\alpha\) is a wiggly arc.
    In this case, \(H^j(X) \neq 0\) for exactly one \(j\), such that \(X \cong H^j(X)[-j]\).
  \item If \(\alpha\) is a general curve, each indecomposable summand of \(H^j(X)\) for any \(j\) is a spherical indecomposable object of \(\mathcal{A}_m\) that arises from a wiggly arc, up to shift.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Let \(X\) be a complex built from an arbitrary (not necessarily wiggly) curve \(\alpha\).
  The results of~\cite{kho.sei:02} imply that \(X\) is spherical and indecomposable.
  Let us now analyse the indecomposable summands of its cohomology objects.

@@

For a general object \(X\) of \(\mathcal{C}_m\), the indecomposable factors of the cohomology filtration can be arbitrary.
However, it turns out that the cohomology filtration is highly constrained if \(X\) is spherical, and these constraints are given by wiggly pseudotriangulations.
\begin{proposition}\label{prop:hn-filtration}
  Let \(X \in \mathcal{C}_{n+1}\) be a spherical object.
  Consider its cohomology filtration
  \[
    \begin{tikzcd}[column sep=0.3cm]
      0 \arrow{rr} && X_{-l} \arrow{dl} \arrow{rr} && \cdots \arrow{rr} \arrow{dl} && X_{-k} = X \arrow{dl}\\
      & H^l(X)[-l] \arrow[dashed]{ul}{+1} && H^{l+1}(X)[-l-1] \arrow[dashed]{ul}{+1} & \cdots & H^k(X)[-k] \arrow[dashed]{ul}{+1}
    \end{tikzcd}
  \]
  Consider the set of all possible indecomposable summands of the cohomology objects \(\{H^j(X)\}\) taken without multiplicity.
  Let \(T\) be corresponding set of wiggly arcs, as explained in~\cref{prop:cohomology-pieces}.
  We have the following results.
  \begin{enumerate}
  \item 
    \(T\) is a subset of a wiggly pseudotriangulation.
  \item \(T\) contains at most one of the two irrelevant arcs.
  \end{enumerate}
\end{proposition}
\begin{proof}
  The proof of this proposition is an easy consequence of analysing the cohomology filtration from the point of view of curves on the disk with marked points.
  Let \(X_1, \ldots, X_k\) be the objects in \(T\) without repetition, which arise from wiggly arcs \(\alpha_1, \ldots, \alpha_k\) as explained in the proof of~\cref{prop:cohomology-pieces}.
  Let \(\beta = (i,j,A,B)\) and \(\beta' = (i',j', A',B')\) be two distinct wiggly arcs among \(\{\alpha_1, \ldots,\alpha_k\}\).
  We check that \(\beta\) and \(\beta'\) are compatible in the sense of~\cref{def:compatible}.

  Recall that \(\beta\) and \(\beta'\) are two pieces of the curve \(\alpha\).
  Therefore \(\beta\) and \(\beta'\) cannot cross, because~\(\alpha\) does not cross itself.
  Suppose that \(\beta\) and \(\beta'\) are non-pointed, and without loss of generality, suppose that \(i' = j\).
  This means that the piece \(\beta\) was cut from \(\alpha\) either because \(i' = j\) is an endpoint of \(\alpha\), or because \(\alpha\) loops around the right of the point \(i'\).
  Moreover, if \(i'\) is an endpoint of \(\alpha\), then the portion of \(\alpha\) closest to \(i'\) lies to the left of \(i'\).

@@

  We have four possible combinations:
  \begin{enumerate}
  \item \(i' = j\) is an endpoint such that \(\alpha\) goes left starting at \(i'\), and \(i' = j\) is an endpoint such that \(\alpha\) goes right starting at \(j\);
  \item \(i' = j\) is an endpoint such that \(\alpha\) goes left starting at \(i'\), and \(\alpha\) loops around the left of the point \(j\);
  \item \(\alpha\) loops around the right of the point \(i' = j\), and \(i' = j\) is an endpoint such that \(\alpha\) goes right starting at \(j\); 
  \item \(\alpha\) loops around the right of the point \(i' = j\), and \(\alpha\) loops around the left of the point \(j\).
  \end{enumerate}
  It is clear that none of these possibilities can occur within the curve \(\alpha\), and hence \(\beta\) and \(\beta'\) must be pointed.
  Therefore \(T = \{\alpha_1, \ldots, \alpha_k\}\) is a subset of a wiggly pseudotriangulation.

@@

However, the grading convention in~\cite{kho.sei:02} is not the same as ours --- our grading is by path length, while the grading in~\cite{kho.sei:02} is one for arrows in one direction and zero for arrows in the opposite direction.
So we must modify their argument slightly.
In particular, the table in~\cite[Lemma 3.20]{kho.sei:02} that computes the bigraded intersection numbers will change, and so will the table in~\cite[Lemma 4.12]{kho.sei:02} that computes the Poincar{\'e} polynomials of \(\Hom(P_k,-)\).
  \asilata{TODO: this is very tedious; how much to write it?}
\end{proof}
\begin{remark}
  \cref{prop:ext1} is very similar to the condition of being compatible in a cluster-tilting object of a cluster category (see, \eg~\cite[Cor.~4.3]{bua.mar.rei.ea:06}) or in a $\tau$-tilting object (see, \eg~\cite{AdachiIyamaReiten}).
  The condition in terms of crossings of curves already appears for some specializations (see, \eg~\cite{FominShapiroThurston, PaluPilaudPlamondon-surfaces}).
  However, in \cref{prop:ext1}, it depends on shifting the objects~\(X\) and~\(X'\) so that they both lie in the standard heart, which is an auxiliary step that is not required in cluster-tilting and $\tau$-tilting theory.
\end{remark}
\begin{remark}
  As stated,~\cref{prop:ext1} in combination with~\cref{prop:hn-filtration} is not expected to be true in greater generality.
  For instance, for the zigzag algebra of Dynkin type~\(D_4\), one can find two indecomposable spherical objects with a non-trivial \(\Ext^1\) that nevertheless appear together in the cohomology filtration of another spherical object.
  However, we expect to see similar constraints on compatibility, which we hope to turn to in forthcoming work.
\end{remark}
\begin{question}
  Suppose we have a collection of spherical objects in the standard heart \(\mathcal{A}_m\) that satisfy the compatibility condition of~\cref{prop:ext1}.
  That is, the \(\Ext^1\) between any pair of objects in this collection is zero.
  \begin{enumerate}
  \item Can we prove properties about a maximal such collection without reference to curves?
  \item In particular, can we understand the combinatorics of maximal such collections without reference to curves?
  \item Given a maximal compatible collection of objects, can we always find a larger spherical object such that the factors of the cohomology filtration of that object are precisely shifts of the objects in the given collection?
  \end{enumerate}
\end{question}

@@

% Local Variables:
% TeX-command-extra-options: "-pvc"
% End:
